### 2015-05-14 Body/LitReview/LitReview.Rnw
Edits 1.3-1.6

The research done in Chapter 2 of this dissertation is a designed experiment to compare the learning outcomes of students receiving traditional and simulation-based curricula for teaching inference in an introductory statistics course.  The term \textit{traditional} refers to a curriculum that introduces statistical inference through the probability distributions and application of mathematical theory in empirical scenarios. This approach is characterized by the use of distributional approximations, cumulative probability tables and formulas for confidence intervals and test statistics. In contrast, the \textit{simulation-based} approach to teaching inference relies on methodology driven by resampling and permutation instead of theoretical probability distributions. This allows for the concepts of statistical inference to be learned through methods such as bootstrap confidence intervals and randomization testing of hypotheses. The following subsections will outline the existing literature on simulation-based inference curricula and their comparison to traditional inference curricula, then explore existing comparative studies within statistics education and lastly explore assessment tools designed for measuring student learning outcomes.

@@

George Cobb is widely regarded as the progenitor of the recent call to use simulation-based methods for teaching introductory statistics. Cobb argued that calculation and graphics were automated and streamlined through an expansion of technology in the curriculum during the statistics education reform movement but nothing was done to change how inference concepts were being taught. Computation has become so cheap and widely available that simulation-based methods that rely heavily on repeated simulation are now becoming practically feasible. He also argues that the traditional theory-based inference methods continue being taught, not because they are better, but because historical momentum drives to their perpetuation. He uses the shift from a earth centered to a sun centered view of the solar system as a metaphor for the the shift from tradition to simulation-based methods for teaching inference; it is not only simpler but it is more accurate to the core concepts of inference \citep{Cobb2007}.

Since 2007 a growing number of educators have joined Cobb in advocating for teaching simulation-based methods in introductory statistics courses. There are a few established groups of authors producing course materials to support the wave of educators adopting the simulation-based approach to inference. \textit{Statistics: Unlocking the Power of Data} is a textbook that follows a traditional progression through data collection, summarization and graphical topics, then uses randomization tests and bootstrap confidence intervals to teach the concepts of inference \citep{Lock5}. The \textit{Introduction to Statistical Investigation} textbook approaches the restructure of the entire introductory course by using the ``spiral approach'' to introduce statistical inference from the very beginning of the course, then adds complexity and related ideas as students repeatedly revisit the inference process \citep{ISI}. Although not in a textbook format, the \textit{Change Agents for Teaching and Learning Statistics (CATALST)} group has a large online collection of freely available lesson plans and course materials that use the simulation-based approach to teaching inference \citep{CATALST}.

@@

There are arguments to be made against the use of experiments in educational studies. \citet{Cook2002} argues that experimental logic strongly supports causal conclusions but outlines several practical reasons why he believes they are not used in education.  He argues that there are often insurmountable problems with properly implementing experiments in schools, results tend to trade external validity for causal isolation and that often observational studies may establish strong enough associations to make the necessary policy decisions. There are also those who are more philosophically opposed to experiments in education. \citet{Howe2004} argues that the quantitative methods used in experimentation are not capable of measuring student learning in any meaningful way and that qualitative methods are the only appropriate approach to assess learning. Aside from his astoundingly imprecise and inept criticism of the experimental methodologies, I fundamentally disagree with his assertion that we cannot quantify understanding or learning. Educators -- perhaps imperfectly -- regularly quantify student knowledge through the grading of assessments and substantial research has been done on tools designed to measure student learning in statistics.

@@

In the experiment comparing simulation-based and traditional inference curricula we need a valid metric for student learning outcomes. There exist a number of tools developed to measure students learning of introductory statistics concepts, some specific to statistical inference. The \textit{Comprehensive Assessment of Outcomes in a first Statistics course (CAOS)} test is a general purpose test that evaluates student understanding in a wide range of introductory statistics topics \citep{DelMas2007}. The \textit{Reasoning about P-values and Statistical Significance (RPASS)} scale was developed to assess student understanding of statistical concepts surrounding hypothesis testing \citep{LaneGetaz2013}. The \textit{Assessment Resource Tools for Improving Statistical Thinking (ARTIST)} project which produced the \textit{CAOS} test also produced several banks of topic specific questions for testing understanding of particular concepts in introductory statistics. These ARTIST question sets were developed through an iterative process using the \textit{Context, Input, Process, Product (CIPP)} evaluation model to improve the quality of learning assessment \citep{Ooms2008}. The intent of our study is to compare the learning outcomes specific to hypothesis testing and confidence intervals, so we elected to use the ARTIST questions sets for these two topics as measures on which to make comparisons of the statistical inference curricula.

@@

Chapter 3 of this dissertation details the construction and evaluation stages in the development of the Shiny Database Sampler; a point-and-click web-based software tool for selecting random subsamples from large databases. The Shiny Database Sampler was created to help fill the void in data technologies discussed in Subsection~\ref{DataTech}, connecting introductory statistics students to available -- but difficult to access -- large data sources. This section highlights pertinent literature, software and data sources used in the development of the Shiny Database Sampler.

The goal of the tool is to allow students to treat large databases as populations from which to collect random samples through specified sampling schemes. The software also provides basic tools to numerically and visually assess the sampled data prior to download. \texttt{R} \citep{R} was chosen as the computational engine to power the Shiny Database Sampler. \texttt{R} provides all of the sampling and computational functionality necessary many of the simple data operations, but two additional packages were integral to the construction of the web interface and database querying. The \texttt{shiny} package \citep{shiny} allows an \texttt{R} session running on a web server to dynamically generate the JavaScript for the user interface accessed via web browser. \texttt{shiny} operates on the principle of reactive programming, where certain changes to the user interface (editing fields, clicking buttons, etc.) are programmed to trigger \texttt{R} to reprocess data objects which are then incorporated back into the JavaScript sent back to the browser for display. This allows for the strength of \texttt{R} for statistical computation to be applied on the server side of a point-and-click web interface. This strength is leveraged in the primary task of collecting random subsamples from a \texttt{MySQL} database \citep{MySQL}. The accompanying query language to interacting with the databases is not simple; certainly no simple enough to expect proficiency from an introductory statistics student. The \texttt{RMySQL} package \citep{RMySQL} allows for database queries to be executed within an \texttt{R} session. By coupling the functionality of \texttt{shiny} and \texttt{RMySQL} we are able to bypass the skill threshold, by allowing the student to set query specifications for random sampling through simplified interface which automates the query behind the scenes.

@@

It is also important to consider \textit{acquiescence}, or a bias toward positive responses, that occurs in self-report data because of a human bias toward the social desirability of affirmation \citep{Furnham1986}. Negatively phrasing half of the items for each latent characteristic, then reversing the scoring, can be used in an attempt to counter-balance the bias. However this double negation may create items that contra-intuitive as measures of positive responses \citep{Friborg2006}. The language of each item must be carefully constructed to attempt to navigate the inherent difficulties with eliciting honest and accurate human responses.

Many tools and concepts were leveraged in the development of the Shiny Database Sampler. Quality characteristics from software engineering, cognitive load theory from educational psychology and human centered design from human computer interaction were fundamental in the construction of the software using \texttt{R}, \texttt{shiny} and \texttt{RMySQL}. Survey methodology used in formal evaluation of the Shiny Database Sampler relies on item response theory.

@@

Methodology related to bivariate data displays and binned data graphics are relevant to work with binned scatterplots. The scatterplot is a widely used tool for display of bivariate quantitative data, that uses points on a Cartesian plane to visually display the coordinates of each pair. \citet{friendly2005early} discuss the development of the scatterplot and state,``Indeed, among all the forms of statistical graphics, the humble scatterplot may be considered the most versatile, polymorphic, and generally useful invention in the entire history of statistical graphics.'' They find that many area plots and time series plots arrived in the eighteenth century \citep{playfair}, but early forms of the scatterplot did not emerge until the nineteenth century \citep{Herschel1833}.

The scatterplot is very effective at visualizing the bivariate distribution of data pairs, but it faces a fundamental challenge; scalability \citep{Theus2006}. While the points do not have any area mathematically, they are visually displayed with round glyphs that occupy space on the Cartesian plane. \textit{Over-plotting} occurs in scatterplots when two or more points share overlapping graphical space, thus obscuring part of the visual information. This problem becomes more pronounced as the number of points increases, substantially lowering the utility of the traditional scatterplot in ever growing modern data sizes. Many adaptations for the scatterplot have been suggested to overcome this problem, most either use changes to the glyphs representing each point (\citealt{tukey},\citealt{few2008solutions}, \citealt{Keim2010GenScatter}, \citealt{Hao2010visual}, \citealt{Janetzko2013Ellipse}) or through plotting binned aggregations of points (\citealt{sunflowerplots}, \citealt{Carr1987}, \citealt{Wickham2013Bin}, \citealt{Liu2013imMens}).

Adaptations that alter the individual points may help in cases of mild over-plotting, but plots that employ binning have superior scalability for visualizing truly massive data. \citet{Liu2013imMens} state that``Visualizing every data point can lead to over-plotting and may overwhelm users' perceptual and cognitive capacities. On the other hand, reducing the data through sampling or filtering can elide interesting structures or outliers.'' Binned scatterplots are a aggregation based plot that is the two-dimensional analog to a histogram, which bins bivariate data and displays frequency through the shade of geometric tiles, as opposed to the height of a bar. Binned scatterplots are constructed by defining a two-dimensional tessellated grid over the range of the data, binning points based on the grid boundaries, then shading matching geometric tiles based on the frequency of points in the corresponding bin. Rectangular binned scatterplots are often referred to as \textit{heatmaps}, which can be traced back to the end of the turn of the twentieth century \citep{Wilkinson2000}.

@@

\citet{scott1992} explains that the histogram convey frequency and relative frequency which are the essence of a density function. Similarly, the binned scatterplot can be thought of a visual estimate of the bivariate density. The binning and rendering of the plot encode the density information into the tiles through shading, therefore work on perception of color is important for understanding how that information is recovered visually. Several authors have tested graphical perception of colors in designed experiments. \citet{cleveland1984graphical} find that shading and color saturation are less accurate graphical aesthetics than position, size and angle. In repeating Cleveland and McGill's experiment using updated technology, \citet{heer2010crowdsourcing} similarly rate color as less precise for representing quantitative information than position, size and angle. \citet{demiralp2014learning} find that people tend to judge color similarity primarily based on hue and secondarily based on shade. \citet{healey1999large} present evidence to suggest that up to seven distinct hues are simultaneously distinguishable for the average person, but there is more difficulty discerning different shades of the same hue. This all points to the difficulty that is faced in extracting frequency information based on the tile shades in a binned scatterplot, therefore special consideration is paid in Chapter 4 to mapping frequency information to shade.

@@

The review of literature identifies important themes, knowledge, and wisdom to be incorporated in the research that follows. The literature on educational technologies has a distinct take-away for work in statistics education. The recursive message resonates clearly; technology presents the opportunity to create a rich learning environment and engage students on a deeper level, however the development and implementation of technology in education requires proper pedagogy and integration into curricula to effectively improve learning. This theme echoes throughout education literature from statistics, and also in the work of our peers in other similar STEM disciplines. Great care must be taken when developing or implementing new educational technologies to ensure that students attentions are guided to learning material, not distracted by the technology. These principles lend well to the pursuit of work on simulation-based learning and the construction of the Shiny Database Sampler. In statistical graphics, the literature commonly reminds us that we not only need to develop efficient ways to visualize information, but we also must critically assess the perceptual ability to accurately recover that information.

>>>>

The research done in Chapter 2 of this dissertation is a designed experiment to compare the learning outcomes of students receiving traditional and simulation-based curricula for teaching statistical inference in an introductory statistics course.  The term \textit{traditional} refers to a curriculum that introduces statistical inference through probability distributions and application of mathematical theory in empirical scenarios. This approach is characterized by the use of distributional approximations, cumulative probability tables and formulas for confidence intervals and test statistics. In contrast, the \textit{simulation-based} approach to teaching statistical inference relies on methodology driven by resampling and permutation instead of theoretical probability distributions. This allows for the concepts of statistical inference to be learned through methods such as bootstrap confidence intervals and randomization testing of hypotheses. The following subsections will outline the existing literature on simulation-based inference curricula and their comparison to traditional inference curricula, then explore comparative studies within statistics education and lastly discuss assessment tools designed for measuring student learning outcomes.

@@

George Cobb is widely regarded as the progenitor of the recent call to use simulation-based methods for teaching introductory statistics. Cobb argued that calculation and graphics were automated and streamlined through an expansion in the use of technology during the statistics education reform movement but nothing was done to change how inference concepts were being taught. Computation has become so cheap and widely available that simulation-based methods that rely heavily on repeated simulation are now becoming practically feasible. He also argues that the traditional theory-based inference methods continue being taught, not because they are better, but because historical momentum drives to their perpetuation. He uses the shift from a earth centered to a sun centered view of the solar system as a metaphor for the the shift from tradition to simulation-based methods for teaching inference; it is not only simpler but it is more accurate to the core concepts of inference \citep{Cobb2007}.

Since 2007 a growing number of educators have joined Cobb in advocating for teaching simulation-based methods in introductory statistics courses. There are a few established groups of authors producing course materials to support the wave of educators adopting the simulation-based approach to inference. \textit{Statistics: Unlocking the Power of Data} is a textbook that follows a traditional progression through data collection, summarization and graphical topics, then uses randomization tests and bootstrap confidence intervals to teach the concepts of statistical inference \citep{Lock5}. The \textit{Introduction to Statistical Investigation} textbook approaches the restructure of the entire introductory course by using the ``spiral approach'' to introduce statistical inference from the very beginning of the course, then adds complexity and related ideas as students repeatedly revisit the inference process \citep{ISI}. Although not in a textbook format, the \textit{Change Agents for Teaching and Learning Statistics (CATALST)} group has a large online collection of freely available lesson plans and course materials that use the simulation-based approach to teaching inference \citep{CATALST}.

@@

There are arguments to be made against the use of experiments in educational studies. \citet{Cook2002} argues that experimental logic strongly supports causal conclusions but outlines several practical reasons why he believes they are not used in education.  He argues that there are often insurmountable problems with properly implementing experiments in schools, results tend to trade external validity for causal isolation, and that often observational studies may establish strong enough associations to make the necessary policy decisions. There are also those who are more philosophically opposed to experiments in education. \citet{Howe2004} argues that the quantitative methods used in experimentation are not capable of measuring student learning in any meaningful way and that qualitative methods are the only appropriate approach to assess learning. Aside from his astoundingly imprecise and inept criticism of the experimental methodologies, his assertion that we cannot quantify understanding or learning is fundmentally incorrect. Educators -- perhaps imperfectly -- regularly quantify student knowledge through the grading of assessments and substantial research has been done on tools designed to measure student learning in statistics.

@@

In the experiment comparing simulation-based and traditional statistics inference curricula we need a valid metric for student learning outcomes. There exist a number of tools developed to measure students learning of introductory statistics concepts, some specific to statistical inference. The \textit{Comprehensive Assessment of Outcomes in a first Statistics course (CAOS)} test is a general purpose test that evaluates student understanding in a wide range of introductory statistics topics \citep{DelMas2007}. The \textit{Reasoning about P-values and Statistical Significance (RPASS)} scale was developed to assess student understanding of statistical concepts surrounding hypothesis testing \citep{LaneGetaz2013}. The \textit{Assessment Resource Tools for Improving Statistical Thinking (ARTIST)} project which produced the \textit{CAOS} test also produced several banks of topic specific questions for testing understanding of particular concepts in introductory statistics. These ARTIST question sets were developed through an iterative process using the \textit{Context, Input, Process, Product (CIPP)} evaluation model to improve the quality of learning assessment \citep{Ooms2008}. The intent of our study is to compare the learning outcomes specific to hypothesis testing and confidence intervals, so we elected to use the ARTIST questions sets for these two topics as measures on which to make comparisons of learning outcomes from the statistical inference curricula.

@@

Chapter 3 of this dissertation details the construction and evaluation stages in the development of the Shiny Database Sampler; a point-and-click web-based software tool for selecting random subsamples from large databases. The Shiny Database Sampler was created to help fill the void in data technologies discussed in Subsection~\ref{DataTech}, connecting introductory statistics students to available -- but difficult to access -- large data sources. This section highlights pertinent literature, software, and data sources used in the development of the Shiny Database Sampler.

The goal of the tool is to allow students to treat large databases as populations from which to collect random samples through specified sampling schemes. The software also provides basic tools to numerically and visually assess the sampled data prior to download. \texttt{R} \citep{R} was chosen as the computational engine to power the Shiny Database Sampler. \texttt{R} provides all of the sampling and computational functionality necessary many of the simple data operations, but two additional packages were integral to the construction of the web interface and database querying. The \texttt{shiny} package \citep{shiny} allows an \texttt{R} session running on a web server to dynamically generate the JavaScript for the user interface accessed via web browser. \texttt{shiny} operates on the principle of reactive programming, where certain changes to the user interface (editing fields, clicking buttons, etc.) are programmed to trigger \texttt{R} to reprocess data objects which are then incorporated into the JavaScript sent back to the browser for display. This allows for the strength of \texttt{R} for statistical computation to be applied on the server side of a point-and-click web interface. This strength is leveraged in the primary task of collecting random subsamples from a \texttt{MySQL} database \citep{MySQL}. The accompanying query language to interacting with the databases is not simple; certainly no simple enough to expect proficiency from an introductory statistics student. The \texttt{RMySQL} package \citep{RMySQL} allows for database queries to be executed within an \texttt{R} session. By coupling the functionality of \texttt{shiny} and \texttt{RMySQL} we are able to bypass the skill threshold, by allowing the student to set query specifications for random sampling through a simplified interface which automates the query behind the scenes.

@@

It is also important to consider \textit{acquiescence}, or a bias toward positive responses, that occurs in self-report data because of a human bias toward the social desirability of affirmation \citep{Furnham1986}. Negatively phrasing half of the items for each latent characteristic, then reversing the scoring, can be used in an attempt to counter-balance this bias. However the resulting double negation may create items that contra-intuitive as measures of positive responses \citep{Friborg2006}. The language of each item must be carefully constructed to attempt to navigate the inherent difficulties with eliciting honest and accurate human responses.

Many tools and concepts were leveraged in the development of the Shiny Database Sampler. Quality characteristics from software engineering, cognitive load theory from educational psychology and human centered design from human computer interaction were fundamental in the construction of the software using \texttt{R}, \texttt{shiny} and \texttt{RMySQL}. Item response theory from statistical survey methodology was used in formal evaluation of the Shiny Database Sampler.

@@

Methodology related to bivariate data displays and binned data graphics are relevant to work with binned scatterplots. The scatterplot is a widely used tool for display of bivariate quantitative data, that uses points on a Cartesian plane to visually display the coordinates of each pair. \citet{friendly2005early} discuss the development of the scatterplot and state, ``Indeed, among all the forms of statistical graphics, the humble scatterplot may be considered the most versatile, polymorphic, and generally useful invention in the entire history of statistical graphics.'' They find that many area plots and time series plots arrived in the eighteenth century \citep{playfair}, but early forms of the scatterplot did not emerge until the nineteenth century \citep{Herschel1833}.

The scatterplot is very effective at visualizing the bivariate distribution of data pairs, but it often fails to overcome a fundamental challenge; scalability \citep{Theus2006}. While the points do not have any area mathematically, they are visually displayed with round glyphs that occupy space on the graphic. \textit{Over-plotting} occurs in scatterplots when two or more points share overlapping graphical space, thus obscuring part of the visual information. This problem becomes more pronounced as the number of points increases, substantially lowering the utility of the traditional scatterplot in ever growing modern data sizes. Many adaptations for the scatterplot have been suggested to overcome this problem, most either use changes to the glyphs representing each point (\citealt{tukey},\citealt{few2008solutions}, \citealt{Keim2010GenScatter}, \citealt{Hao2010visual}, \citealt{Janetzko2013Ellipse}) or through plotting binned aggregations of points (\citealt{sunflowerplots}, \citealt{Carr1987}, \citealt{Wickham2013Bin}, \citealt{Liu2013imMens}).

Adaptations that alter the individual points may help in cases of mild over-plotting, but plots that employ binning have superior scalability for visualizing truly massive data. \citet{Liu2013imMens} state that ``Visualizing every data point can lead to over-plotting and may overwhelm users' perceptual and cognitive capacities. On the other hand, reducing the data through sampling or filtering can elide interesting structures or outliers.'' The binned scatterplot is an aggregation based plot that is the two-dimensional analog to a histogram, which bins bivariate data and displays frequency through the shade of geometric tiles, as opposed to the height of a bar. Binned scatterplots are constructed by defining a two-dimensional tessellated grid over the range of the data, binning points based on the grid boundaries, then shading matching geometric tiles based on the frequency of points in the corresponding bin. Rectangular binned scatterplots are often referred to as \textit{heatmaps}, which can be traced back to the end of the turn of the twentieth century \citep{Wilkinson2000}.

@@

\citet{scott1992} explains that the histogram convey frequency and relative frequency which are the essence of a density function. Similarly, the binned scatterplot can be thought of a visual estimate of the bivariate density. The binning and rendering of the plot encode the density information into the tiles through shading, therefore research on perception of color is important for understanding how that information is recovered visually. Several authors have tested graphical perception of colors in designed experiments. \citet{cleveland1984graphical} find that shading and color saturation are less accurate graphical aesthetics than position, size and angle. In repeating Cleveland and McGill's experiment using updated technology, \citet{heer2010crowdsourcing} similarly rate color as less precise for representing quantitative information than position, size and angle. \citet{demiralp2014learning} find that people tend to judge color similarity primarily based on hue and secondarily based on shade. \citet{healey1999large} present evidence to suggest that up to seven distinct hues are simultaneously distinguishable for the average person, but there is more difficulty discerning different shades of the same hue. This all points to the difficulty that is faced in extracting frequency information based on the tile shades in a binned scatterplot, therefore special consideration is paid in Chapter 4 to mapping frequency information to shade.

@@

The review of literature identifies important themes, knowledge, and wisdom to be incorporated in the research that follows. The literature on educational technologies has a distinct take-away for work in statistics education. The recursive message resonates clearly; technology presents the opportunity to create a rich learning environment and engage students on a deeper level, however the development and implementation of technology in education requires proper pedagogy and integration into curricula to effectively improve learning. This theme echoes throughout education literature from statistics, and also in the work of our peers in similar STEM disciplines. Great care must be taken when developing or implementing new educational technologies to ensure that students attentions are guided to learning new concepts, not distracted by the technology. These principles lend well to the pursuit of work on simulation-based learning and the construction of the Shiny Database Sampler. In statistical graphics, the literature commonly reminds us that we not only need to develop efficient ways to visualize information, but we also must critically assess the perceptual ability to accurately recover that information.
